# Introduction {.unnumbered}

```{r}
#| label: setup
#| warning: false
#| message: false

source(here::here("scripts/setup.R"))

sf_use_s2(FALSE)

```

```{r}
#| label: load-port-data
##| warning: false
##| message: false

ports_normalized <- read_csv(here("data/ports-enriched4.csv"),
                             show_col_types = FALSE) %>%
  mutate( across(.cols = c(port2, level2, country), ~str_trim(str_to_lower(.x)) ) ) |>
  arrange(port2, country, level2) |>
  fill(country, .direction = "down") |>
  distinct(port2, country, level2, direction, port, .keep_all = TRUE)
port_lon_lat <- read_csv(here("data/port_lon_lat.csv"),
                         show_col_types = FALSE) %>%
  mutate( across(.cols = c(port2, level2, country), ~str_trim(str_to_lower(.x)) ) ) %>%
  filter(!is.na(port2))
df_ports <- left_join(ports_normalized, port_lon_lat, by = "port2") %>%
  mutate_if(is.character, ~map_chr(.x, iconv, "UTF-8", "UTF-8", sub='')) %>% # just in case
  mutate(level2 = coalesce(level2.x, level2.y),
         country = coalesce(country.x, country.y)) %>%
  select(-ends_with(c(".x", ".y")))

```

```{r}
#| warning: false
#| message: false

source(here("scripts/load-voyages.R"))

```

Two hundred and fifty years ago sailings ships were the engines of colonization and commerce. Where did they sail? How far and how
long? We can gain some insight by making use of the Climatological Database for the World's Oceans (CLIWOC), which includes 287K
observations from logbooks of `r nrow(count(all_voyages, ShipName))` ships from `r nrow(count(all_voyages, Nationality))` countries,
mostly British, Dutch, French, and Spanish) that made `r nrow(count(all_voyages, ShipName, VoyageIni))` journeys. The digitized
records are mostly from journeys between 1750 and 1815. These journey during this portion of the [Age of
Sail](https://en.wikipedia.org/wiki/Age_of_Sail) influenced (and sometimes directly motivated) by these historical events:

-   [Seven years war](https://en.wikipedia.org/wiki/Seven_Years%27_War) (1756-1763) which included the [French and Indian
    War](https://en.wikipedia.org/wiki/French_and_Indian_War) in North America
-   [American War of Independence](https://en.wikipedia.org/wiki/American_Revolutionary_War), [French
    Revolution](https://en.wikipedia.org/wiki/French_Revolution) and resulting [French Revolutionary
    Wars](https://en.wikipedia.org/wiki/French_Revolutionary_Wars), and others wars of independence in the Americas, including
    [Haiti](https://en.wikipedia.org/wiki/Haitian_Revolution), and the [Spanish American
    colonies](https://en.wikipedia.org/wiki/Spanish_American_wars_of_independence) (the latter continued beyond 1815 for many
    colonial lands)
-   Voyages of discovery, including two by [James Cook](https://en.wikipedia.org/wiki/James_Cook)
-   [Battle of Waterloo](https://en.wikipedia.org/wiki/Battle_of_Waterloo) and [Congress of
    Vienna](https://en.wikipedia.org/wiki/Congress_of_Vienna), bringing to end the [Napoleonic
    Wars](https://en.wikipedia.org/wiki/Napoleonic_Wars)
-   The waxing preeminence of the British on the worlds' seas.

<br>

This data set only hints at the human drama: the strict and sometimes shockingly excessive punishment, the high mortality rate, the
deprivations, and the near-unimaginable sufferings inherent in the slave trade. Or the triumphs of winning a hard-fought, costly
battle, capturing an enemy ship, and, through many perils, bringing a fortune into one's home port. Historical fiction brings this
period to life. I recommend Patrick O'Brian's Aubrey and Maturin series starting with [Master and
Commander](https://en.wikipedia.org/wiki/Master_and_Commander). Or start with the movie [Master and Commander: The Far Side of the
World](https://en.wikipedia.org/wiki/Master_and_Commander:_The_Far_Side_of_the_World), which includes events from a number of the
books.

<br>
