# Voyage maps

```{r}
#| label: setup
#| warning: false
#| message: false

source(here::here("scripts/setup.R"))

sf_use_s2(FALSE)

```

```{r}
#| label: load-port-data

ports_normalized <- read_csv(here("data/ports-enriched4.csv"),
                             show_col_types = FALSE) %>%
  mutate( across(.cols = c(port2, level2, country), ~str_trim(str_to_lower(.x)) ) ) |>
  arrange(port2, country, level2) |>
  fill(country, .direction = "down") |>
  distinct(port2, country, level2, direction, port, .keep_all = TRUE)
port_lon_lat <- read_csv(here("data/port_lon_lat.csv"),
                         show_col_types = FALSE) %>%
  mutate( across(.cols = c(port2, level2, country), ~str_trim(str_to_lower(.x)) ) ) %>%
  filter(!is.na(port2))
df_ports <- left_join(ports_normalized, port_lon_lat, by = "port2") %>%
  mutate_if(is.character, ~map_chr(.x, iconv, "UTF-8", "UTF-8", sub='')) %>% # just in case
  mutate(level2 = coalesce(level2.x, level2.y),
         country = coalesce(country.x, country.y)) %>%
  select(-ends_with(c(".x", ".y")))

```

```{r}
#| label: create-world-and-water

# World map example from https://www.simoncoulombe.com/2020/11/animated-ships/
  
world <- rnaturalearth::ne_countries(scale='medium',returnclass = 'sf') %>%
  st_transform(crs = my_proj)

# create water polygon for background 
lats <- c(90:-90, -90:90, 90)
longs <- c(rep(c(180, -180), each = 181), 180)
water_outline <- 
  list(cbind(longs, lats)) %>%
  st_polygon() %>%
  st_sfc(crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs") %>% 
  st_sf() %>%
  st_transform(crs = my_proj)

```

```{r}
#| warning: false
#| message: false

source(here("scripts/load-voyages.R"))

source(here("scripts/plot-voyages-functions.R"))

```

## Voyages to or from Virginia ("Hampton Road")

British ships made most of the voyages to Virginia in this data set; Dutch ships made a few. We see a pattern in the routes between
the Mid-Atlantic coast and Northern Europe, due to the [prevailing winds](https://en.wikipedia.org/wiki/Prevailing_winds): when
leaving Europe, south near the Azores or Canary Islands then west across the Atlantic; when leaving North America: a more northerly
path.

For tabular data see [Hampton Road]

<br>

```{r}
#| label: fig-hamton-road
#| fig-cap: "Voyages to or from Hampton Roads, Virginia"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

voyage_list <- df_voyages %>% 
               filter(port_from == "hampton road" | port_to == "hampton road")

plot_voyages(ships = voyage_list$ShipName,
             from = "hampton road",
             to = "hampton road"
)
```

<br>

Looking at the same voyages by year, we see nearly the whole time period represented.

<br>

```{r}
#| label: fig-hamton-road-by-year
#| fig-cap: "Voyages to or from Hampton Roads, Virginia, by year"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(ships = voyage_list$ShipName,
             from = "hampton road",
             to = "hampton road",
             color_var = "year"
)

```

<br>

## Voyages between US and Netherlands

Our data set includes Dutch voyages from Boston, New York, and Virginia. All occurred after US independence--the first one directly after.

Also see table at [US and Netherlands].

<br>

```{r between-us-netherlands, out.width="100%", echo = TRUE}
#| label: between-us-netherlands
#| fig-cap: "Voyages between the U.S. and Netherlands"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

ship_list <- 
  c(df_voyages %>% 
      filter(country_from == "united states" & country_to == "netherlands") %>%
      pull(ShipName),
    df_voyages %>%
      filter(country_from == "netherlands" & country_to == "united states") %>%
      pull(ShipName)
  )

plot_voyages(ships = ship_list)
plot_voyages(ships = ship_list,
             color_var = "year")

```

<br>

## Voyages between US and France

These voyages occurred in the final months of the American Revolutionary War in 1783 and afterwards.

<br>

```{r}
#| label: between-us-france
#| fig-cap: "Voyages between the U.S. and France"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

ship_list <- 
  c(df_voyages %>% 
      filter(country_from == "united states" & country_to == "france") %>%
      pull(ShipName),
    df_voyages %>%
      filter(country_from == "france" & country_to == "united states") %>%
      pull(ShipName)
  )

plot_voyages(ships = ship_list)

plot_voyages(ships = ship_list,
             color_var = "year")

```

<br>

## Voyages between Canada and France

Most French voyages occurred before or during the [Seven Years War](https://en.wikipedia.org/wiki/Seven_Years%27_War), after which France's territorial holdings in North America were limited to the small islands of [St Pierre and
Miquelon](https://en.wikipedia.org/wiki/Saint_Pierre_and_Miquelon). They remain a part of France today.

See also table [Canada and France]

<br>

```{r}
#| label: between-canada-france
#| fig-cap: "Voyages between Canada and France"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

ship_list <- 
  c(df_voyages %>% 
      filter(country_from == "canada" & country_to == "france") %>%
      pull(ShipName),
    df_voyages %>%
      filter(country_from == "france" & country_to == "canada") %>%
      pull(ShipName)
  )

plot_voyages(ships = ship_list,
             color_var = "year")

```

<br>

## Voyages to or from Florida

[St. Augustine](https://en.wikipedia.org/wiki/St._Augustine,_Florida) was founded in 1565 by Spanish explorers and has been
inhabited continuously since then--the longest of any city in the continental US. It became the capital of British East Florida in
1763, returned to Spanish control in 1783, then Spain ceded it to the USA in 1819. Our data set includes only British voyages.

Also see table [Florida].

<br>

```{r}
#| label: to-from-florida
#| fig-cap: "Voyages to or from Florida"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(
  from = filter(df_ports, level2 == "florida") %>% pull(port2),
  to = filter(df_ports, level2 == "florida") %>% pull(port2),
  color_var = "year"
)

```

<br>

```{r}

```

## All voyages of ships that went to to or from Florida at least once

Looking at all the voyages in our data set undertaken by ships that traveled from or to Florida, we see a much wider web going as
far as India, Ceylon (now Sri Lanka), Malaysia, and the Dutch East Indies (now Indonesia and nearby island nations). These
`r nrow(distinct(voyage_list, ShipName))` ships were part of the British Royal Navy.

<br>

```{r florida2-year, out.width="100%", echo = TRUE}
#| label: to-from-florida-all-ships
#| fig-cap: "All voyages of ships that went to or from Florida at least once"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

voyage_list <- df_voyages %>% 
               filter(str_detect(port_to, "florida") | str_detect(port_from, "florida"))

plot_voyages(ships = voyage_list$ShipName,
             color_var = "year")

```

<br>

## Veracruz, Mexico

The primary port in the Spanish colony of Mexico was Veracruz. Cuba was settled earlier and remained a regional center in this time
period.

<br>

```{r}
#| label: mexico-eastcoast
#| fig-cap: "All voyages to or from Veracruz, Mexico"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(to = "veracruz")
plot_voyages(to = "veracruz", color_var = "year")
plot_voyages(from = "veracruz")
plot_voyages(from = "veracruz", color_var = "year")

```

## Explorations of the northwest coast of North America

Our data set includes Spanish, British and French voyages of discovery along the western coast of North America. The Spanish went
further south; the British, further north.

-   The historic [La Boussole](https://en.wikipedia.org/wiki/French_ship_Boussole_(1781)) ("the compass") captained by
    [Jean-François de Galaup, comte de Lapérouse](https://en.wikipedia.org/wiki/Jean-François_de_Galaup,_comte_de_Lapérouse), which
    disappeared after leaving Botany Bay in 1788.
-   The first [HMS Queen Charlotte](https://en.wikipedia.org/wiki/HMS_Queen_Charlotte), launched in 1790 and sank in 1800.
-   The Spanish
    [Aranzazu](https://text-message.blogs.archives.gov/2017/10/19/establishing-and-disbanding-the-neah-bay-settlement-1792/) which
    was looking for the rumored northwest passage. For more background to Spanish aims, activities, and rivalry with the British in
    the American Northwest, see
    [here](https://text-message.blogs.archives.gov/2017/10/12/225-years-ago-spanish-explorations-of-the-pacific-northwest-and-the-first-spanish-settlement-in-washington-state-nunez-gaona-neah-bay-1792-part-ii-spanish-explorations-of-the-pacific-northwest-a/).

See also table [Northwest America]

<br>

```{r}
#| label: nw-north-america
#| fig-cap: "All voyages to or from the US Coastal Northwest"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

nw_explorers <- 
  get_voyage_details(range_lon = c(-170, -100),
                     range_lat = c(15, 90)
                     )

plot_voyages(ships = nw_explorers$ShipName,
             range_lon = c(-170, -100),
             range_lat = c(15, 90)
)

plot_voyages(ships = nw_explorers$ShipName,
             range_lon = c(-170, -100),
             range_lat = c(15, 90),
             color_var = "year")

```

<br>

## James Cook's first and second voyages of discovery

Our data set includes portions of [James Cook's](https://en.wikipedia.org/wiki/James_Cook) first circumnavigation westward for the
purpose of observing the [1769 transit of Venus in Tahiti](https://en.wikipedia.org/wiki/1769_transit_of_Venus_observed_from_Tahiti)
and then searching for unknown land to the south, [Terra Astralis, Incognita](https://en.wikipedia.org/wiki/Terra_Australis),
believed to exist further south than Australia, which was then called New Holland. On this journey he landed in Australia at a place
he named [Botany Bay](https://en.wikipedia.org/wiki/Botany_Bay) in present-day Sydney. Our records include portions of Cook's second
voyage from Britain eastward to New Zealand and eventually around the South American cape and back to Britain. He never found
Antarctica, despite coming close, and his experience convinced people that Terra Astralis, Incognita did not exist.

See also table [James Cook's journeys]

<br>

```{r}
#| label: cook
#| fig-cap: "James Cook's voyages in the data set"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

ship_list <- c("ENDEAVOUR", "RESOLUTION")
cooks_first_journey <- c("1768-01-01", "1771-12-31")
cooks_second_journey <- c("1772-01-01", "1775-12-31")

plot_voyages(ships = ship_list,
             range_dates = cooks_first_journey,
             color_var = "year"
)

plot_voyages(ships = ship_list,
             range_dates = cooks_second_journey,
             color_var = "year"
)

```

## Maps: South America

## To and or from Brazil, Uruguay, Argentina

Most voyages in our data from or to South America we between to [Montevideo](https://en.wikipedia.org/wiki/Montevideo) and Spanish
ports. Montevideo was Spain' primary naval base in this period. The trade from the interior via the [Río de la
Plata](https://en.wikipedia.org/wiki/R%C3%ADo_de_la_Plata) (River Plate) passed through Montevideo or [Buenos
Aires](https://en.wikipedia.org/wiki/Buenos_Aires), which sits on the opposite coast of the estuary.

At least some of the British voyages likely contributed to the [Battle of Montevideo
(1807)](https://en.wikipedia.org/wiki/Battle_of_Montevideo_(1807)) or its aftermath, when the British captured and held Montevideo
for much of that year.

The voyages to or from Spain occurred before or during the 1780s--well before British blockades of [Spanish
ports](https://en.wikipedia.org/wiki/Cádiz) 1797-1893 and 1803-1808.

To:

<br>

```{r}
#| label: spanish-america
#| fig-cap: "All voyages to or from Spanish America"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(to = df_voyages %>% 
      filter(country_to %in% c("brazil", "uruguay", "argentina" ))  %>%
      pull(port_to)
      )

plot_voyages(to = df_voyages %>% 
      filter(country_to %in% c("brazil", "uruguay", "argentina" ))  %>%
      pull(port_to),
      color_var = "year"
      )

```

<br>

From:

<br>

```{r}
#| label: from-spanish-america
#| fig-cap: "All voyages from Spanish America"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(from = df_voyages %>% 
      filter(country_from %in% c("brazil", "uruguay", "argentina" ))  %>%
      pull(port_from)
      )

plot_voyages(from = df_voyages %>% 
      filter(country_from %in% c("brazil", "uruguay", "argentina" ))  %>%
      pull(port_from),
      color_var = "year"
      )

```

<br>

## The challenge of rounding the Horn traveling west

The track of the french ship La Boussole illustrates the difficulty [square-rigged](https://en.wikipedia.org/wiki/Square_rig) ships
faced from storms and contrary winds.

<br>

```{r}
#| label: around-horn-la-boussole
#| fig-cap: "La Boussole's track around the horn of South America"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(from = df_voyages %>% 
      filter(country_from %in% c("brazil"),
             Nationality == "FRENCH")  %>%
      pull(port_from),
      ships = "LA BOUSSOLE"
      )

```

```{r}
#| label: whole-voyage-la-boussole
#| fig-cap: "La Boussole's track around the horn of South America"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(
      ships = "LA BOUSSOLE"
      )

```

<br>

## Maps: African Guinea

## Triangular trade

```{r}
guinea_coast <- c("senegal", "guinea-bissau", "guinea", "sierra leone", "liberia", "ivory coast", "ghana", "togo", "benin", "nigeria", "cameroon", "equatorial guinea", "gabon")

```

In our data set: Dutch and British voyages before 1780 to [Guinea](https://en.wikipedia.org/wiki/Guinea_(region)), modern day
`r glue_collapse(str_to_title(guinea_coast), sep = ", ", last = " and ")`. Most voyages were to/from slave trading posts or colonial
fortresses. We see two parts of the [triangular trade](https://en.wikipedia.org/wiki/Triangular_trade) route: from Europe to Guinea,
then to the Americas.

<br>

```{r}
#| label: to-guinea-coast
#| fig-cap: "Voyages to the Guinea coast"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(to = df_voyages %>% 
      filter(country_to %in% guinea_coast)  %>%
      pull(port_to),
      range_lat = c(-35, 20)
      )

plot_voyages(to = df_voyages %>% 
      filter(country_to %in% guinea_coast)  %>%
      pull(port_to),
      range_lat = c(-35, 20),
      color_var = "year"
      )

```

<br>

```{r}
#| label: from-guinea-coast
#| fig-cap: "Voyages from the Guinea coast"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(from = df_voyages %>% 
      filter(country_from %in% guinea_coast)  %>%
      pull(port_from),
      range_lat = c(-35, 20)
      )

plot_voyages(from = df_voyages %>% 
      filter(country_from %in% guinea_coast)  %>%
      pull(port_from),
      range_lat = c(-35, 20),
      color_var = "year"
      )

```

<br>

## Maps: India and Indonesia

## To and from Penang, Malaysia

Our data set includes 28 voyages to or from Penang. There are just enough to see patterns in the routes: leaving the UK for
Malaysia, ships stayed near Africa, veered toward Brazil, then picked up the prevailing westerlies beyond 35 degrees South, perhaps
taking advantage of the [Antarctic Circumpolar Current](https://en.wikipedia.org/wiki/Antarctic_Circumpolar_Current) as well.

<br>

```{r}
#| label: to-penang
#| fig-cap: "Voyages to Penang, Malaya"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(to = "penang")

```

<br>

In contrast, when traveling the other direction, captains minimized the time they faced the prevailing westerlies, just barely
rounding the Cape, then traveling much further into the north Atlantic before turning towards the UK.

<br>

```{r}
#| label: from-penang
#| fig-cap: "Voyages from Penang, Malaya"
#| fig-height: 8
#| fig-width: 10
#| column: page-right
plot_voyages(from = "penang")

```

<br>

## To and from India

This pattern is clearer still the tracks to and from India.

<br>

```{r}
#| label: to-india
#| fig-cap: "Voyages to India"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(to = df_voyages %>%
               filter(country_to == "india") %>%
               pull(port_to)
)

plot_voyages(to = df_voyages %>%
               filter(country_to == "india") %>%
               pull(port_to),
             color_var = "year"
)

```

<br>

From India

<br>

```{r from-india, out.width="100%", echo = TRUE}
#| label: from-india
#| fig-cap: "Voyages from India"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(from = df_voyages %>%
               filter(country_from == "india") %>%
               pull(port_from)
)

```

<br>

## To and from Indonesia

The same pattern is largely true for ships going to and from Indonesia. Most were Dutch; the main difference in their route was
staying near 40 degrees South longer on the way to Indonesia.

<br>

```{r}
#| label: to-indonesia
#| fig-cap: "Voyages to Indonesia"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(to = df_voyages %>%
               filter(country_to == "indonesia") %>%
               pull(port_to)
)

plot_voyages(to = df_voyages %>%
               filter(country_to == "indonesia") %>%
               pull(port_to),
             color_var = "year"
)

```

<br>

From Indonesia

<br>

```{r}
#| label: from-indonesia
#| fig-cap: "Voyages from Indonesia"
#| fig-height: 8
#| fig-width: 10
#| column: page-right


plot_voyages(from = df_voyages %>%
               filter(country_from == "indonesia") %>%
               pull(port_from)
)

plot_voyages(from = df_voyages %>%
               filter(country_from == "indonesia") %>%
               pull(port_from),
             color_var = "year"
)

```

<br>

## To and from China (Hong Kong and Macau)

This data set doesn't have many voyages to or from China.

<br>

```{r}
#| label: to-china
#| fig-cap: "Voyages to China"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(to = df_voyages %>% filter(country_to == "china") %>% pull(port_to))
plot_voyages(to = df_voyages %>% filter(country_to == "china") %>% pull(port_to), color_var = "year")

```

```{r}
#| label: from-china
#| fig-cap: "Voyages from China"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

plot_voyages(from = df_voyages %>% filter(country_from == "china") %>% pull(port_from))
plot_voyages(from = df_voyages %>% filter(country_from == "china") %>% pull(port_from), color_var = "year")

```

<br>

## The Importance of Chronometers

Perhaps you've noticed the unlikely number of horizontal tracks, for example, above when leaving Jakarta and traveling to western
India or Yemen or ships approaching St Helena from the East. Ships would round the Cape and find the right latitude well to the east
of St Helena. Then they could sail directly west to the island. Why? Because it's hard to find a small island in a big ocean, and
until there were sufficiently accurate chronometers, it was much easier to calculate an accurate latitude than longitude. In the
period of this data set, chronometer technology improved a lot, and one can see in the trend towards lesser safety margin in later
years.

<br>

```{r}
#| label: chronometers-st-helena
#| fig-cap: "As chronometers got better, captains could assume more accurate logitude calculations"
#| fig-height: 8
#| fig-width: 10
#| column: page-right

voyage_list <- df_voyages %>%
               filter(country_from == "india" & port_to == "st helena")

plot_voyages(ships = voyage_list$ShipName,
             from = voyage_list$port_from,
             to = "st helena", 
             range_lon = c(-10, 60),
             range_lat = c(-44, -5),
             color_var = "year"
)

```

<br>
