# Voyage tables

```{r}
#| label: setup
#| warning: false
#| message: false

source(here::here("scripts/setup.R"))

source(here("scripts/load-voyages.R"))

source(here("scripts/plot-voyages-functions.R"))

```

Below are selected data tables corresponding to some of the maps plotted earlier.

## Hampton Roads {#sec-hampton-roads}

Voyages to/from Hampton Roads ("hampton road") in table form. Column names are a subset of the variables in the data frame I used
for plotting. Ones with capital letters were in the original data set; I added the ones in lower case.

<br>

```{r}
#| label: tbl-hampton-road-by-year-table
#| tbl-cap: "Voyages to or from Hampton Roads, Virginia"
#| fig-height: 8
#| column: page-right

voyage_list <- df_voyages %>% 
               filter(port_from == "hampton road" | port_to == "hampton road")

summarize_voyages(ships = voyage_list$ShipName,
                  from = "hampton road",
                  to = "hampton road"
)

```

<br>

## US and Netherlands {#sec-us-netherlands}

```{r}
#| label: tbl-between-us-netherlands
#| tbl-cap: "Voyages between the USA and Netherlands"
#| column: page-right

ship_list <- 
  c(df_voyages %>% 
      filter(country_from == "united states" & country_to == "netherlands") %>%
      pull(ShipName),
    df_voyages %>%
      filter(country_from == "netherlands" & country_to == "united states") %>%
      pull(ShipName)
  )

summarize_voyages(ships = ship_list)

```

<br>

## Canada and France {#sec-canada-france}

```{r}
#| label: tbl-between-canada-france
#| tbl-cap: "Voyages between Canada and France"
#| column: page-right

ship_list <- 
  c(df_voyages %>% 
      filter(country_from == "canada" & country_to == "france") %>%
      pull(ShipName),
    df_voyages %>%
      filter(country_from == "france" & country_to == "canada") %>%
      pull(ShipName)
  )

summarize_voyages(ships = ship_list)

```

\<br

## Florida {#sec-florida}

```{r}
#| label: tbl-to-from-Florida
#| tbl-cap: "Voyages to or from Florida"
#| column: page-right

voyage_list <- df_voyages %>% 
  filter(str_detect(port_to, "florida") | str_detect(port_from, "florida"))

summarize_voyages(ships = voyage_list$ShipName,
                  from = c("florida", "cape florida", "florida bay"),
                  to = c("florida", "cape florida", "florida bay")
)

```

<br>

### American Northwest {#sec-american-northwest}

```{r }
#| label: tbl-american-nw
#| tbl-cap: "Voyages to or from the American Northwest"
#| column: page-right

nw_explorers <- 
  get_voyage_details(range_lon = c(-170, -100),
                     range_lat = c(15, 90)
                     )

summarize_voyages(ships = nw_explorers$ShipName,
                  range_lon = c(-170, -100),
                  range_lat = c(15, 90),
)

```

<br>

## James Cook's journeys {#sec-cook}

```{r}
#| label: tbl-cook
#| tbl-cap: "James Cook's Voyages in the data set"
#| column: page-right

ship_list <- c("ENDEAVOUR", "RESOLUTION")
cooks_journeys <- c("1768-01-01", "1774-12-31")

summarize_voyages(ships = ship_list,
                  range_dates = cooks_journeys
)

```

<br>